Efficient simulation of individual-based population models

### 2024-03-27 article.qmd
update

\begin{algorithm}
\caption{IBM simulation algorithm (without events of Poissonian intensity)}
\begin{algorithmic}
    \State \textbf{Input:} Initial population $Z_0$, horizon $T > 0$, and events described by:
        \State - Intensity functions and bounds $(\lambda^e, \bar \lambda^e)$ for $e \in \mathcal{E}$ and $(W^e, \bar W^e)$ for $e \in \mathcal{E}_W$ 
        \State - Event action functions $\phi^e(t, I)$ for $e \in \mathcal{E} \cup \mathcal{E}_W$ (see @tbl-event-action)
    \State \textbf{Output:} Population $Z_T$
    \State Initialization $T_0 \longleftarrow 0$, $\bar T_0 \longleftarrow 0$
    \While{$T_k < T$}
        \Repeat
            \State increment iterative variable $\ell \longleftarrow \ell+1$
            \State compute next proposed time $\bar T_{\ell} \longleftarrow \bar T_{\ell-1} + \mathcal{E}\big(\bar \Lambda(N_{T_k}) \big)$
            \State draw a proposed event $\bar E_\ell \sim \mathcal{U}\{p_e\}$ with $p_e = \frac{\bar \lambda^e \mathbf{1}_{e \in \mathcal{E}} + \bar W^e N_{T_k} \mathbf{1}_{e \in \mathcal{E}_W}}{\sum_{e \in \mathcal{E}} \bar \lambda^e + \sum_{e \in \mathcal{E}_W} \bar W^e N_{T_k}}$
            \State draw an individual index $\bar K_\ell \sim \mathcal{U}(\{1,\dots,N_{T_k}\})$ 
            \State draw $\bar \Theta_\ell \sim \mathcal{U}([0, \bar \lambda^{\bar E_\ell}])$ if $\bar E_\ell \in \mathcal{E}$ or $\bar \Theta_\ell \sim \mathcal{U}([0, \bar W^{\bar E_\ell} N_{T_k}])$ if $\bar E_\ell \in \mathcal{E}_W$
        \Until{accepted event $\bar \Theta_\ell \le \lambda^{\bar E_\ell}_{\bar T_\ell}(I_{\bar K_\ell},Z_{T_{k}})$}
        \State increment iterative variable $k \longleftarrow k+1$ 
        \State record $(T_k, E_k, I_k) \longleftarrow (\bar T_{\ell}, \bar E_\ell,  I_{\bar K_\ell})$ as accepted time, event and individual
        \State update the population $Z_{T_k} = Z_{T_{k-1}} + \phi^{E_k}(T_k, I_k)$
    \EndWhile
\end{algorithmic}
\end{algorithm}
```

@@

#### Event creation with individual intensity {.unnumbered}

@@

#### Event creation with interaction intensity {.unnumbered}

@@

#### Events creation with Poisson and Inhomogeneous Poisson intensity {.unnumbered}

@@

For Poisson (constant) intensities the events are created with the function
```{R}
#| eval: false
```

The following example creates a death event, where individuals die at a constant intensity `lambda` (which has to be in the list of model parameters):

@@

This section provides an example of how to use the `IBMPopSim` package
to simulate a heterogeneous life insurance portfolio (see
also `vignette(IBMPopSim_insurance_portfolio)`).

We consider an insurance portfolio consisting of male policyholders, of
age greater than 65. These policyholders are characterized by their age,
assumed to be less than $a_{max} = 110$, and risk class
$x\in \mathcal X =\{1,2\}$.

@@

**Death events** A baseline age and time specific death rate is first
calibrated on "England and Wales (EW)" males mortality historic
data[^1], and projected for 30 years using the Lee-Carter model with the
package `StMoMo` (see [@stmomo]). The forecasted baseline death
intensity is denoted by $d(t,a)$, defined by:
$$
d(t,a) = \sum_{k=0}^{29}\mathbf{1}_{\{k\leq t < k+1\}} d_k(a), \quad \forall \; t\in [0,30] \text{ and } a \in [65, a_{max}],
$$ {#eq-insurance-baseline}
with $d_k(a)$ the point estimate of the forecasted mortality rate for
age $a$ and year $k$.\
Individuals in risk class 1 are assumed to have mortality rates that are
20% higher than the baseline mortality (for instance, the risk class
could refer to smokers), while individuals in risk class 2 are assumed
to have mortality rates that are 20% lower than the baseline (non
smokers). The death intensity of an individual $I= (\tau_b, \infty, x)$,
of age $a(I,t) = t - \tau_b$ at time $t$ and in risk class
$x \in \{1, 2\}$ is thus the function $$\label{eq-insurance-deathrates}
\lambda^d(t,I) = \alpha_x d(t,a(I,t)), \quad \alpha_1 = 1.2, \quad \alpha_2 = 0.8.$$
In particular, the death intensity verifies Assumption TODO
[2](#AssumptionIntensity1){reference-type="ref"
reference="AssumptionIntensity1"} since:
$$
\lambda^d(t,I) \leq \bar d : = \alpha_1 \sup_{t \in [0,30]} d(t,a_{max}).
$$ {#eq-insurance-bound-deathrates}

>>>>

\begin{algorithm}
\caption{IBM simulation algorithm (without events of Poissonian intensity)}
\begin{algorithmic}
    \State \textbf{Input:} Initial population $Z_0$, horizon $T > 0$, and events described by:
        \State - Intensity functions and bounds $(\lambda^e, \bar \lambda^e)$ for $e \in \mathcal{E}$ and $(W^e, \bar W^e)$ for $e \in \mathcal{E}_W$ 
        \State - Event action functions $\phi^e(t, I)$ for $e \in \mathcal{E} \cup \mathcal{E}_W$ 
    \State \textbf{Output:} Population $Z_T$
    \State Initialization $T_0 \longleftarrow 0$, $\bar T_0 \longleftarrow 0$
    \While{$T_k < T$}
        \Repeat
            \State increment iterative variable $\ell \longleftarrow \ell+1$
            \State compute next proposed time $\bar T_{\ell} \longleftarrow \bar T_{\ell-1} + \mathcal{E}\big(\bar \Lambda(N_{T_k}) \big)$
            \State draw a proposed event $\bar E_\ell \sim \mathcal{U}\{p_e\}$ with $p_e = \frac{\bar \lambda^e \mathbf{1}_{e \in \mathcal{E}} + \bar W^e N_{T_k} \mathbf{1}_{e \in \mathcal{E}_W}}{\sum_{e \in \mathcal{E}} \bar \lambda^e + \sum_{e \in \mathcal{E}_W} \bar W^e N_{T_k}}$
            \State draw an individual index $\bar K_\ell \sim \mathcal{U}(\{1,\dots,N_{T_k}\})$ 
            \State draw $\bar \Theta_\ell \sim \mathcal{U}([0, \bar \lambda^{\bar E_\ell}])$ if $\bar E_\ell \in \mathcal{E}$ or $\bar \Theta_\ell \sim \mathcal{U}([0, \bar W^{\bar E_\ell} N_{T_k}])$ if $\bar E_\ell \in \mathcal{E}_W$
        \Until{accepted event $\bar \Theta_\ell \le \lambda^{\bar E_\ell}_{\bar T_\ell}(I_{\bar K_\ell},Z_{T_{k}})$}
        \State increment iterative variable $k \longleftarrow k+1$ 
        \State record $(T_k, E_k, I_k) \longleftarrow (\bar T_{\ell}, \bar E_\ell,  I_{\bar K_\ell})$ as accepted time, event and individual
        \State update the population $Z_{T_k} = Z_{T_{k-1}} + \phi^{E_k}(T_k, I_k)$
    \EndWhile
\end{algorithmic}
\end{algorithm}
```

@@

_Event creation with individual intensity_

@@

_Event creation with interaction intensity_

@@

_Events creation with Poisson and Inhomogeneous Poisson intensity_

@@

For Poisson (constant) intensities the events are created with the function
```{r}
#| eval: false
mk_event_poisson(type = "TYPE",
                 name = "NAME",
                 intensity = "CONSTANT", ...)
```

The following example creates a death event, where individuals die at a constant intensity `lambda` (which has to be in the list of model parameters):
```{r}
#| eval: false
mk_event_poisson(type = "death,
                 intensity = "lambda")
```

When the intensity $(\mu^e(t))$ depends on time, the event can be created similarly by using the function
```{r}
#| eval: false
mk_event_inhomogeneous_poisson(type = "TYPE",
                               name = "NAME",intensity = "CONSTANT", ...)
```

### Event kernel code {#sec-event-kernel-code}

When an event occurs, the events kernels $k^e$ specify how the event modifies the population. The events kernels are defined in the `kernel_code` parameter of the `mk_event_CLASS(type = "TYPE", name ="NAME", ...)` function. The `kernel_code` is `NULL` by default and doesn't have to be specified for death, exit events and birth events, but mandatory for entry and swap events. Recall that the `kernel_code` argument is a string composed of a few lines of `C++` code, characterizing the individual characteristics following the event.
@tbl-events-variables summarizes the list of available variables that can be used in the `kernel_code`.

- **Death/Exit event** If the user defines a death event, the death date of the current individual `I` is set automatically to the current time `t`. Similarly, when an individual `I` exits the population,`I.out` is set automatically to `TRUE` and his exit time is recorded as a \"death\" date. For these events types, the `kernel_code` doesn't have to be specified by the user.

- **Birth event** The default generated event kernel is that an individual `I` gives birth to a new individual `newI` of age 0 at the current time `t`, with same characteristics than the parent `I`. If no kernel is specified, the default generated `C++` code for a birth event is:
```{r}
#| eval: false
individual newI = I;
newI.birth_date = t;
pop.add(newI);
```
The user can modify the birth kernel, by specify the argument `kernel_code` of `mk_event_CLASS`. In this case, the generated code is
```{r}
#| eval: false
individual newI = I;
newI.birth_date = t;
_KERNEL_CODE_
pop.add(newI);
```
where `_KERNEL_CODE_` is replaced by the content of the `kernel_code` argument.

- **Entry event** When an individual `I` enters the population, `I.entry` is set automatically as the date at which the individual enters the population. When an entry occurs the individual entering the population is not of age $0$. In this case, the user must specify the `kernel_code` argument indicating how the age and characteristics of the new individual are chosen. For instance, the code below creates an event of type `entry`, named `ev_example`, where individuals enter the population at a Poisson constant intensity. When an individual `newI` enters the population at time `t`, his age is chosen as a normally distributed random variable, with mean 20 and variance 4.
```{r}
#| eval: false
mk_event_poisson(
    type = "entry", 
    name = "ev_example", 
    intensity = "lambda", 
    kernel_code = "
        double a_I = max(CNorm(20, 2), 0.); 
        newI.set_age(a_I, t);
    ")
```

- **Swap event** The user must specify the `kernel_code` argument indicating how the characteristics of an individual are modified following a swap.

::: {#tbl-events-variables}

Variable           Description
------------------ -------------------------------------------------------------
Variable           Description
`I`                Current individual
`J`                Another individual in the population (only for interaction)
`t`                Current time
`pop`              Current population (vector)
`newI`             Available only for birth and entry events.   
Model parameters   Depends on the model

`C++` variables available for events kernel code
:::

::: {#rem}
When there are several events of the same type, the user can identify which events generated a particular event by adding a characteristic to the population recording the event name/id when it occurs. See e.g. `vignette(IBMPopSim_human_pop)` for an example with different causes of death.
:::

## Model creation {#sec-model-creation}

Once the population, the events, and model parameters are defined, the IBM model is created using the function `mk_model`.

```{r}
#| eval: false
model <- mk_model(characteristics = get_characteristics(pop_init),
                  event = events_list,
                  parameters = model_params)
```
During this step which can take a few seconds, the model is created and compiled using the `Rcpp` package. The model structure in `IBMPopSim` is that the model depends only on the population characteristics' and parameters names and types, rather than their values. This means that once the model has been created, various simulations can be done with different initial populations and different parameters values.

::: {#exm-model-creation}
Here is an example of model with a population structured by age and gender, with birth and death events. The death intensity of an individual of age $a$ is $d(a) = \alpha \exp(\beta a),$ and females between 15 and 40 can give birth with birth intensity $b(a) = \bar \lambda^b \mathbf{1}_{[15,40]}.$ The newborn is a male with probability $p_{male}$.
```{r}
#| eval: false
params <- list("p_male"= 0.51,
               "birth_rate" = stepfun(c(15,40),c(0,0.05,0)),
               "death_rate" = gompertz(0.008,0.02))

death_event <- mk_event_individual(type = "death", name= "my_death_event",
                  intensity_code = "result = death_rate(age(I,t));")

birth_event <- mk_event_individual( type = "birth",
                  intensity_code = "if (I.male)
                                        result = 0;
                                    else
                                        result=birth_rate(age(I,t));",
                  kernel_code = "newI.male = CUnif(0, 1) < p_male;")
pop <- population(EW_pop_14$sample)

model <- mk_model(characteristics = get_characteristics(pop),
                  events = list(death_event,birth_event),
                  parameters = params)
```
:::

## Simulation {#sec-package-simulation}

The simulation of the IBM is based on the algorithms presented in @sec-simulation-algo and @sec-simulation-randomized.
The user has first to specify bounds for the intensity or interaction functions of each event type. The random evolution of the population can then be simulated over a period of time $[0,T]$ by calling the function `popsim`.

### Events bounds

Since the IBM simulation algorithm is based on an acceptance-rejection method for simulating random times, the user has to specify bounds for the intensity (or interaction) functions of each event (see @cnj-intensity1 and @cnj-intensity2).
These bounds should be stored in a named vector, where for event $e$, the name corresponding to the event bound $\bar{\mu}^e$, $\bar{\lambda}^e$ or $\bar{W}^e$ is the event `name` defined during the event creation step.

In @exm-model-creation from previous section the intensity bound for birth events is $\bar\lambda_b$. Since the death intensity function is not bounded, the user will have to specify a maximum age $a_{\max}$ in `popsim` (all individuals above $a_{\max}$ die automatically). Then, the bound for death events is $\bar \lambda_d = \alpha\exp(\beta a_{\max}).$ 
In the example, the death event has been named `my_death_event`. No name has been specified for the birth event which thus has the default name `birth`. Then,
```{r}
#| eval: false
a_max <- 120 # maximum age
events_bounds <- c("my_death_event" = params$death_rate(a_max),
                   "birth" = max(params$birth_rate))
```

Once the model and events bounds have been defined, a random trajectory of the population can be simulated by calling
```{r}
#| eval: false
sim_out <- popsim(model, pop, events_bounds, params,
                  age_max = a_max, time = 30)
```

_Optional parameters_

@@

This section provides an example of how to use the `IBMPopSim` package to simulate a heterogeneous life insurance portfolio (see also [`vignette(IBMPopSim_insurance_portfolio)`](https://daphnegiorgi.github.io/IBMPopSim/articles/IBMPopSim_insurance_portfolio.html)).

We consider an insurance portfolio consisting of male policyholders, of age greater than 65. These policyholders are characterized by their age, assumed to be less than $a_{\max} = 110$, and risk class $x \in \mathcal X = \{1, 2\}$.

@@

**Death events** A baseline age and time specific death rate is first
calibrated on "England and Wales (EW)" males mortality historic
data[^1], and projected for 30 years using the Lee-Carter model with the
package `StMoMo` (see [@stmomo]). The forecasted baseline death
intensity is denoted by $d(t,a)$, defined by:
$$
d(t,a) = \sum_{k=0}^{29}\mathbf{1}_{\{k\leq t < k+1\}} d_k(a), \quad \forall \; t\in [0,30] \text{ and } a \in [65, a_{\max}],
$$ {#eq-insurance-baseline}
with $d_k(a)$ the point estimate of the forecasted mortality rate for
age $a$ and year $k$.\
Individuals in risk class 1 are assumed to have mortality rates that are
20% higher than the baseline mortality (for instance, the risk class
could refer to smokers), while individuals in risk class 2 are assumed
to have mortality rates that are 20% lower than the baseline (non
smokers). The death intensity of an individual $I= (\tau_b, \infty, x)$,
of age $a(I,t) = t - \tau_b$ at time $t$ and in risk class
$x \in \{1, 2\}$ is thus the function $$\label{eq-insurance-deathrates}
\lambda^d(t,I) = \alpha_x d(t,a(I,t)), \quad \alpha_1 = 1.2, \quad \alpha_2 = 0.8.$$
In particular, the death intensity verifies Assumption TODO
[2](#AssumptionIntensity1){reference-type="ref"
reference="AssumptionIntensity1"} since:
$$
\lambda^d(t,I) \leq \bar d : = \alpha_1 \sup_{t \in [0,30]} d(t,a_{\max}).
$$ {#eq-insurance-bound-deathrates}
